3D Regression Heat Map Analysis of Population Study Data

### 2015-06-22 draft.tex
revised cube occurences in the cover letter; updated Section titles
added Bernhards revision of page 1

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subject's lifestyle, medical status as well as medical image data.
%%
Statistical regression analysis is used to evaluate these features and to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set w.r.t. a target feature, no suitable tool is available for epidemiological researchers.

%\com{New Solution}.
%%
We propose an analysis approach of epidemiological data sets by incorporating all features in an exhaustive regression-based analysis.
%%
This approach combines all \emph{independent features} with respect to a \emph{target feature}.
%%
It provides a visualization that reveals insights into the data by highlighting relationships.
%%
The \emph{3D Regression Heat Map}, a novel 3D visual encoding, acts as an overview of the whole data set.
%%
It shows all combinations of two to three independent features w.r.t. a specific target disease.
%%
%A 3D visualization of all combinations of two to three independent features towards a target acts as an overview of the whole data set, the \emph{3D Regression Heat Map}.
%%
Slicing through the \emph{3D Regression Heat Map} allows for the detailed analysis of the underlying relationships.
%%
Expert knowledge about disease-specific hypotheses can be included into the analysis by adjusting the regression model formulas.
%%
Furthermore, the influences of features can be assessed using a difference view comparing different calculation results.
%%
%\com{Validation \& Results}.
%%
We applied our \emph{3D Regression Heat Map} method to a hepatic steatosis data set to reproduce results from a data mining-driven analysis.
%%
A qualitative analysis was conducted on a breast fat data set.
%%
We were able to derive new hypotheses about relations between breast fat density and breast lesions with breast cancer.
%%
%\com{Implications}.
%%
With the \emph{3D Regression Heat Map}, we present a visual overview of epidemiological data that allows for the first time an interactive regression-based analysis of large feature sets with respect to a disease.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analysis, Epidemiology, Breast Cancer, Hepatic Steatosis}

@@

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
Furthermore, the insights can be used to derive recommendations regarding a healthy lifestyle or to provide information about widespread diseases.
%%
During the standard workflow, physicians transform observations into hypotheses.
%%
The hypotheses are depicted using epidemiological features and are then statistically analyzed.

%An important epidemiological tool for deriving such features is the evaluation of a \emph{cohort study}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
An important epidemiological tool for deriving such features are \emph{cohort studies}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
%%
To reduce any selection bias, subjects are randomly invited without a focus on a specific disease.
%%
Hence, a wide range of features is acquired.
%
Social and lifestyle factors, prior or current diseases and medications as well as medical parameters, such as blood pressure, are gathered.
%%
Medical image data from non-radiating modalities, e.g., magnetic resonance imaging (MRI), is also acquired in modern studies.
%%
These data may be quantified based on user-defined landmarks, which describe attributes like the shape, volume or diameter of a structure.
%%

Testing features for association with diseases using regression models is one of the most important epidemiological tools.
%%
Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the higher dimensional problem and the required subject count.
%%
Due to the amount of data and only limited overview techniques, possible correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented in prior work \cite{Klemm2014VIS} are not tailored to a specific target feature.
%%
They mostly highlight correlations between features, which are known to the domain expert (e.g., correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts, into overview visualizations to support a hypothesis-free analysis or an analysis w.r.t. a specific disease or hypothesis.
%%
This is achieved by providing template regression formulas, which are applied to all potential feature combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual features on the process.

@@

\design{
\begin{itemize}
	%%
  \item An overview visualization design based on feedback of epidemiological domain experts to support hypothesis-generation w.r.t. a target feature using regression models.
	%%
  \item Applying domain knowledge in the analysis using freely adjustable regression formulas.
	%%
  \item Selecting suitable metrics for representing and comparing regression models, as well as incorporating additional metrics for details on demand to assess the model quality.
	%%
  \item An openly available web-based system, which incorporates open formats to be applicable for data of other application domains.
	%%
\end{itemize}
}
% \begin{itemize}
% 	%%
% 	\item An overview visualization technique representing feature interactions using target features.
% 	%%
% 	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
% 	%%
% 	\item Freely adjustable regression formulas to provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
% 	%%
% 	\item The analysis of confounding features by providing comparison views between different formula results.
% 	%%
% 	\item The expansion and adaption of our method to any data with our open web-based system.
% 	%%
% \end{itemize}
%%
% \noindent The term `Regression Cube' has been used for the detailed analysis of one or a small number of regression models \cite{Ahmadi, Chan}.
% %%
% They extend 2D scatter plots to display detailed information about a regression model.
% %%
% In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as \emph{Regression Cube}.
% %%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{sec:EpidemiologicalWorkflow}
%%
Epidemiological research is performed by experts from different academic disciplines, such as epidemiologists, physicians, statisticians, and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features with statistical methods.
%%
As described by Thew et al.~\cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{enumerate}
	\item Clinicians/Epidemiologists derive hypotheses via clinical observations, experimental studies or literature research.
	%\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of features depicting the hypothesis and include confounding features.
	\item Statisticians assess the association of the derived features w.r.t the investigated disease.
\end{enumerate}
%%
Relative risks can be determined if a statistical resilient association of features with a condition is extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data regarding the classical sequential epidemiological workflow.
%%
Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

@@

	\textbf{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
%\end{itemize}
%%
The currently selected slice is denoted with a semi-transparent gray plane.
%%
\design{
Early designs rendered the whole \emph{3D Regression Heat Map}, which made it hard to assess the position of the plane.
%%
%Since the regression metrics are mirrored along the diagonal, we decided to use this space to support better identification of the current slice.
Since the regression metrics are mirrored along the diagonal, the space available from visualizing only the prism generated from the upper half of the heat map diagonal is used to display information about the currently selected plane.
}
%%
The \magenta{regression metric values} are projected on this plane to provide an overlapping-free view.
%%
This allows for a easier to identify the current slice.
%%
\paragraph{2D Heat Map Plot Slice Visualization [4.DD].}
%%
The 2D heat map (Fig.~\ref{fig:System}c) shows all values below the matrix diagonal of the current slice.
%%
It creates a optical equivalence with the 3D heat map.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation-based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D heat map.
%%

>>>>

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subject's lifestyle, medical status as well as medical image data.
%%
Statistical regression analysis is used to evaluate these features and to identify feature combinations indicating a disease (the \emph{target feature}).
%%
%Although there is a strong demand for overview visualizations of a whole data set w.r.t. a target feature, no suitable tool is available for epidemiological researchers.

%\com{New Solution}.
%%
We propose an analysis approach of epidemiological data sets by incorporating all features in an exhaustive regression-based analysis.
%%
This approach combines all \emph{independent features} w.r.t. a \emph{target feature}.
%%
It provides a visualization that reveals insights into the data by highlighting relationships.
%%
The \emph{3D Regression Heat Map}, a novel 3D visual encoding, acts as an overview of the whole data set.
%%
It shows all combinations of two to three independent features with a specific target disease.
%%
%A 3D visualization of all combinations of two to three independent features towards a target acts as an overview of the whole data set, the \emph{3D Regression Heat Map}.
%%
Slicing through the \emph{3D Regression Heat Map} allows for the detailed analysis of the underlying relationships.
%%
Expert knowledge about disease-specific hypotheses can be included into the analysis by adjusting the regression model formulas.
%%
Furthermore, the influences of features can be assessed using a difference view comparing different calculation results.
%%
%\com{Validation \& Results}.
%%
We applied our \emph{3D Regression Heat Map} method to a hepatic steatosis data set to reproduce results from a data mining-driven analysis.
%%
A qualitative analysis was conducted on a breast fat data set.
%%
We were able to derive new hypotheses about relations between breast fat density and breast lesions with breast cancer.
%%
%\com{Implications}.
%%
With the \emph{3D Regression Heat Map}, we present a visual overview of epidemiological data that allows for the first time an interactive regression-based analysis of large feature sets with respect to a disease.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analysis, regression, Epidemiology, Breast Cancer, Hepatic Steatosis}

@@

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups \cite{Fletcher2012}.
%%
Furthermore, the insights can be used to derive recommendations regarding a healthy lifestyle or to provide information about widespread diseases.
%%
During the standard workflow, physicians transform observations into hypotheses.
%%
The hypotheses are depicted using epidemiological features and are then statistically analyzed.

%An important epidemiological tool for deriving such features is the evaluation of a \emph{cohort study}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
An important epidemiological tool for deriving such features are \emph{cohort studies}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
%%
To reduce any selection bias, subjects are randomly invited without a focus on a specific disease.
%%
Hence, a wide range of features is acquired.
%
Social and lifestyle factors, prior or current diseases and medications as well as medical parameters, such as blood pressure, are gathered.
%%
%Medical image data from non-radiating modalities, e.g., magnetic resonance imaging (MRI), is also acquired in modern studies.
%%
%These data may be quantified based on user-defined landmarks, which describe attributes like the shape, volume or diameter of a structure.
%%

Testing features for associations with diseases using regression models is one of the most important epidemiological tools.
%%
Using regression analysis to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the higher dimensional problem and the required subject count.
%%
Due to the amount of data and only limited overview visualizations, possible correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented in prior work \cite{Klemm2014VIS} are not tailored to a specific target feature.
%%
They mostly highlight correlations between features, which are known to the domain expert (e.g., correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts, into overview visualizations to support a hypothesis-free analysis or an analysis w.r.t. a specific disease or hypothesis.
%%
This is achieved by providing template regression formulas, which are applied to all potential feature combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual features on the process.

@@

\design{
\begin{itemize}
	%%
  \item An overview visualization design based on feedback of epidemiological domain experts to support hypothesis-generation w.r.t. a target feature using regression models.
	%%
  \item Applying domain knowledge in the analysis using freely adjustable regression formulas.
	%%
  \item Selecting suitable metrics for representing and comparing regression models, as well as incorporating additional metrics for details on demand to assess the model quality.
	%%
  \item An openly available web-based system, which incorporates open formats to be applicable for data of other application domains.
	%%
\end{itemize}
}
% \begin{itemize}
% 	%%
% 	\item An overview visualization technique representing feature interactions using target features.
% 	%%
% 	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
% 	%%
% 	\item Freely adjustable regression formulas to provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
% 	%%
% 	\item The analysis of confounding features by providing comparison views between different formula results.
% 	%%
% 	\item The expansion and adaption of our method to any data with our open web-based system.
% 	%%
% \end{itemize}
%%
% \noindent The term `Regression Cube' has been used for the detailed analysis of one or a small number of regression models \cite{Ahmadi, Chan}.
% %%
% They extend 2D scatter plots to display detailed information about a regression model.
% %%
% In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as \emph{Regression Cube}.
% %%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow, epidemiological data in general and the SHIP in particular.
%%
\subsection{Epidemiological Workflow} \label{sec:EpidemiologicalWorkflow}
%%
Epidemiological research is performed by experts from different academic disciplines, such as epidemiologists, physicians, statisticians, and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features with statistical methods.
%%
As described by Thew et al.~\cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{enumerate}
	\item Clinicians/Epidemiologists derive hypotheses via clinical observations, experimental studies or literature research.
	%\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of features depicting the hypothesis and include confounding features.
	\item Statisticians assess the association of the derived features w.r.t the investigated disease.
\end{enumerate}
%%
Relative risks can be determined if a statistical resilient association of features with a condition is extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data regarding the classical sequential epidemiological workflow.
%%
Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

@@

	\textbf{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
%\end{itemize}
%%
The currently selected slice is denoted with a semi-transparent gray plane.
%%
\design{
Early designs rendered the whole \emph{3D Regression Heat Map}, which made it hard to assess the position of the plane.
%%
%Since the regression metrics are mirrored along the diagonal, we decided to use this space to support better identification of the current slice.
Since the regression metrics are mirrored along the diagonal, the space available from visualizing only the prism generated from the upper half of the heat map diagonal is used to display information about the currently selected plane.
}
%%
The \magenta{regression metric values} are projected on this plane to provide an overlapping-free view.
%%
This allows for a easier to identify the current slice.
%%
\paragraph{2D Heat Map Slice Visualization [4.DD].}
%%
The 2D heat map (Fig.~\ref{fig:System}c) shows all values below the matrix diagonal of the current slice.
%%
It creates a optical equivalence with the 3D heat map.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation-based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D heat map.
%%
